Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

### 2021-08-25 2021/Entropy/entropy.tex
after translation

%=================================================================
% Full title of the paper (Capitalized)
\Title{Ускорение метода глобальной оптимизации за счет выделения локальных экстремумов на основе машинного обучения}
%		   Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Department of Mathematical Software and Supercomputing Technologies, Lobachevsky University, 603950 Nizhni Novgorod, Russia; ilya.lebedev@itmm.unn.ru (I.L.)}
%$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
В статье рассматриваются задачи глобальной оптимизации и численные методы их решения. Задачи указанного класса являются вычислительно-трудоемкими, т.к. целевая функция может являться многоэкстремальной, недифференцируемой, и, как правильно, заданной в виде «черного ящика». В проведенном исследовании использовался детерминированный алгоритм поиска глобального экстремума, который не основан на идеях мультистарта и nature-inspired algorithms. В статье приведены вычислительные правила одномерного алгоритма и nested optimization scheme, которая позволяет применять его для решения сложных многомерных задач. Отметим, что сложность решения задач глобальной оптимизации существенным образом зависит от наличия многих локальных экстремумов. В работе предложен алгоритм выделения областей притяжения локальных минимумов, основанный на использовании методов машинного обучения (в частности, деревьев решений). Использование в выделенных областях алгоритмов локальной оптимизации позволяет значительно ускорить сходимость глобального поиска за счет снижения числа поисковых испытании в окрестностях локальных минимумов. Приведены результаты вычислительных экспериментов, проведенных на нескольких сотнях задач глобальной оптимизации разной размерности, которые подтверждают эффект ускорения сходимости (в терминах числа поисковых испытаний, требующихся для решения задачи с заданной точностью).
}

@@

Успешное применение методов машинного обучения (ML) для решения широкого спектра задач приводит к созданию новых подходов к эффективному использованию идей ML в различных областях приложений. 
Примером класса задач, в которых методы машинного обучения продемонстрировали свою эффективность, являются задачи выявления основных свойств исследуемых явлений (например, физических, экономических или социальных), которые характеризуются стохастической природой или наличием скрытых параметров \cite{Golovenkin2020,Gonoskov2019}.
Методы машинного обучения успешно используются и для решения сложных задач вычислительной математики, например, для simulation of dynamical systems \cite{Seleznev2019}, решения ordinary, partial or stochastic differential equations \cite{Lagaris1998,Blechschmidt2021,Xu2020} .

Одной из таких сложных задач вычислительной математики, к решению которой можно так или иначе применить методы машинного обучения, являются задачи глобальной оптимизации. 
В таких задачах, как правило, не представляется возможным найти решение аналитически и возникает необходимость построения численных методов для его поиска.

Проблема численного решения задач оптимизации сопряжена со значительными трудностями. Во многом они связаны с размерностью и видом целевой функции. При этом наиболее сложными являются задачи, в которых целевая функция является многоэкстремальной, недифференцируемой и, более того, заданной в форме черного ящика (т. е. в виде некоторой вычислительной процедуры, на вход которой подается аргумент, а выходом является соответствующее значение функции). Методы решения именно таких задач рассматриваются в данной статье.

Можно выделить несколько подходов к построению численных методов решения задач глобальной оптимизации. 
%Надо ли добавить алгоритмы на основе метамоделей ?
Ряд алгоритмов основан на идее мультистарта: запуск локального поиска либо начиная с разных стартовых точек, либо с использованием разных параметров. Методы локальной оптимизации обладают большой скоростью сходимости. При этом одной из основных проблем в мультистартовых схемах является выбор начальных точек, которые бы соответствовали областям притяжения различных локальных решений. 
К решению данной проблемы могут быть успешно применены методы машинного обучения. 
Например, в \cite{RinnooyKan1987} для выбора перспективных стартовых точек использовались методы кластерного анализа. 
В \cite{Cassioli2012} область для запуска локального метода выделялась на основе классификации стартовых точке с помощью support vector machine.

Еще одним популярным классом методов, применимым к решению задач глобальной оптимизации, являются метаэвристические алгоритмы. 
Многие из них основаны на имитации процессов, протекающих в живой природе. Для настройки параметров таких алгоритмов также применимы методы машинного обучения. Например, в \cite{Jin2005} приведен обзор использования методов машинного обучения в эволюционных алгоритмах.

Следует отметить, что алгоритмы последних двух классов не обеспечивают гарантированную сходимость к решению задачи и проигрывают детерминированным алгоритмам по качеству работы \cite{Kvasov2018,Sergeyev2018} (e.g., measured by the number of correctly solved problems from a certain set). Поэтому перспективным является использование именно детерминированных методов.

@@

Для алгоритма глобального поиска в разное время было предложено несколько стратегий ускорения его работы (в терминах числа итераций, требующихся для решения задачи с заданной точностью). В данной работе предлагается новый подход к ускорению, основанный на выявления областей притяжения локальных минимумов с использованием методов машинного обучения. Выявление областей притяжения и запуск в этих областях локального поиска позволяет существенным образом сократить число испытаний, требующееся методу для достижения глобальной сходимости. Сказанное подтверждается результатами экспериментов, проведенных на серии из нескольких сотен тестовых задач.

@@

В данной работе мы будем рассматривать задачи глобальной оптимизации вида
\begin{eqnarray}\label{main_problem}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in \text{R}^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}
Задача (\ref{main_problem}) рассматривается в предположении, что целевая функция является многоэкстремальной, задана как ``черный ящик'', а вычисление ее значений связано с решением задачи численного моделирования и является трудоемкой операцией.

Для многих прикладных задач типичной является ситуация, когда ограниченное изменение вектора параметров $y$ вызывает ограниченное изменение значений $\varphi(y)$. Математической моделью, описывающей указанное предположение, является предположение о выполнимости условия Липшица
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty.
\]
Предположение липшицевости целевой функции типично для многих подходов к разработке оптимизационных алгоритмов \cite{Jones1993,Pinter1996,Zilinskas2008,Evtushenko2009}.
При этом многие известные методы основаны на различных способах разбиения области поиска на систему подобластей и последующего выбора наиболее перспективной подобласти для размещения очередного испытания (вычисления значения целевой функции) \cite{Jones2009,Zilinskas2010,Evtushenko2013,Kvasov2013,Paulavicius2016}.

Важным свойством задач глобальной оптимизации является тот факт, что в отличие от задач на поиск локального экстремума, глобальный минимум является интегральной характеристикой решаемой задачи. Чтобы убедиться, что точка $y^*\in D$ является решением задачи, недостаточно исследовать лишь ее окрестность, требуется исследование всей области поиска. Как результат, при минимизации существенно многоэкстремальных функций численный метод должен построить покрытие области поиска и число узлов этого покрытия увеличивается экспоненциально с ростом размерности. 
Эта особенность обуславливает высокую трудоемкость решения задач многоэкстремальной оптимизации, и размерность является решающим фактором, влияющим на сложность их решения.

Для преодоления сложностей, вызванных размерностью решаемой задачи, в многоэкстремальной оптимизации широко используются различные подходы к уменьшению размерности.  
Например, simplicial or diagonal partition области поиска позволяет использовать для решения исходной многомерной задачи методы решения одномерных задач (see, for example, \cite{PaulaviciusZilinskas2014,Sergeyev2017,Sergeyev2013}). 
Известным подходом к редукции размерости является также использование Peano space-filling curves mapping the multidimensional domain onto one-dimensional interval \cite{Strongin2000}.

В данной работе мы будем использовать еще один способ, основанный на the nested optimization scheme \cite{Shi2000,Grishagin2001,VanDam2010,Grishagin2015} and in its generalization \cite{Grishagin2016,Grishagin2016_1}.
The nested optimization scheme, с одной стороны, не ухудшает свойств минимизируемой функции (в отличие от редукции с использованием Peano curves), а с другой стороны, не требует использования сложных структур данных для поддержки симплексных или диагональных разбиений допустимой области. 
Одновременно с этим схема вложенной оптимизации позволяет свести решение многомерной задачи к решению набора стандартных одномерных задач и использовать широкий спектр алгоритмов одномерной глобальной оптимизации для их решения.

@@

В качестве стандартной задачи рассмотрим одномерную задачу многоэкстремальной оптимизации 
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
с целевой функцией, удовлетворяющей условию Липшица.

Приведем описание global search algorithm (GSA) для решения базовой задачи в соответствии с \cite{Strongin2000}.
В процессе своей работы GSA порождает последовательность точек $x^i$, в которых вычисляются значения целевой функции $z^i=\varphi(x^i)$. 
Будем называть процесс вычисления значения целевой функции \textit{trial}.

В соответствии с алгоритмом первые два испытания проводятся в граничных точках отрезка $[0,1]$, т.е. $x^0=0,\;x^1=1$. 
В этих точках вычисляются значения целевой функции $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ и устанавливается значение счетчика $k=1$. 
Точка очередного испытания $x^{k+1}, k\geq 1,$ выбирается в соответствии со следующими действиями.

 Step 1. Перенумеровать нижним индексом (начиная с 0) точки $x^i,\:0\leq i\leq k$, проведенных испытаний в порядке возрастания координаты, т.е.
\begin{equation}\label{xt}
0=x_0<x_1<\ldots <x_{k}=1.
\end{equation} 
Сопоставить точкам $x_i, \; 0\leq i\leq k$, вычисленные в них значения целевой функции $z_i=\varphi(x_i), \; 0\leq i\leq k$.

Step 2. Вычислить максимальное абсолютное значение относительной первой разности
\begin{equation}\label{mu}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i},
\end{equation}
где $\Delta_i = x_i-x_{i-1}$. Если вычисленное в соответствии с данной формулой значение равно нулю, то положить $\mu = 1$.

Step 3. Для всех интервалов $(x_{i-1},x_i),1\leq i\leq k$,  вычислить значение
\begin{equation}\label{R}
R(i)=r\mu\Delta_i+\frac{(z_i-z_{i-1})^2}{r\mu\Delta_i}-2(z_i+z_{i-1}),
\end{equation} 
называемое \textit{характеристикой} интервала; величина $r>1$ является параметром алгоритма.

Step 4. Найти интервал $(x_{t-1},x_t)$ с максимальной характеристикой
\begin{equation}\label{MaxR}
R(t)=\max_{1\leq i\leq {k}}R(i).
\end{equation}
Если максимальная характеристика соответствует нескольким интервалам, то в качестве $t$ выбрать минимальное число, удовлетворяющее (\ref{MaxR}).

Step 5. Провести новое испытание в точке
\begin{equation}\label{xk1}
x^{k+1}=\frac{1}{2}(x_{t-1}+x_t) - \frac{z_t-z_{t-1}}{2r\mu}.
\end{equation}

Алгоритм прекращает свою работу при выполнении условия $\Delta_t<\epsilon$; здесь $t$ из (\ref{MaxR}), а $\epsilon>0$ есть заданная точность. 
В качестве оценки решения задачи выбираются значения 
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i).
\]

Теоретические условия, определяющие сходимость алгоритма, представлены в \cite{Strongin2000}. 
Работа алгоритма при минимизации конкретной многоэкстремальной функции, которая задаются в соответствии с формулой (\ref{hill}), проиллюстрирована на figure \ref{fig1}. При запуске алгоритма использовался параметр $r=2.2$ из (\ref{R}) и значение $\epsilon = 10^{-3}$ в критерии остановки метода.
На рисунке \ref{fig1} изображен график objective function и точки 71 поискового испытания, потребовавшихся GSA для решения задачи с указанной точностью. Наглядно видна проблема всех методов глобальной оптимизации -- концентрация точек в окрестностях локальных минимумов задачи, которые не являются глобальным решением.

\begin{figure}[H]
\includegraphics[width=1.0\linewidth]{HillAGP90.png}
\caption{Минимизация функции вида (\ref{hill}) с помощью GSA.}
\label{fig1}
\end{figure}

\subsection{Machine Learning Regression как инструмент для выделения областей притяжения локальных экстремумов}\label{TreeGSA}

Функции, рассматриваемые в рамках данного исследования, принадлежат классу липшицевых функций. Поэтому классические методы регрессии (например, полиномиальная регрессия, когда функция приближается полиномом заданной степени) не будут хорошо соответствовать поведению функции.

Более мощным инструментом являются регрессионные сплайны. При построении регрессионного сплайна область определения разбивается на $K$ непересекающихся подобластей, в каждой из которых функция аппроксимируется полиномом. При разбиении интервала на достаточное число подобластей позволяет очень точно аппроксимировать исходную функцию.

Построить регрессию можно также и с использованием такого мощного инструмента, как искусственные нейронные сети. 
Для построения регрессии могут быть использованы сети разного типа, например, multilayer perceptron, radial basis function network, и т.д. 
Однако в обоих этих случаях модель (сплайн или же нейронная сеть) сама становится сложной для того анализа, который требуется выполнить в решаемой проблеме (выделение областей притяжения локальных экстремумов).

Поэтому в рамках проведенного исследования для анализа локального поведения функции мы выбрали регрессионную модель, основанную на деревьях решений. 
Например, если целевая функция хорошо аппроксимируется полиномами, то полиномиальная регрессия, конечно же, будет адекватно передавать свойства функции. Однако если между имеет место более сложная зависимость, тогда дерево решений может превзойти по качеству аппроксимации классические варианты регрессии.
%Здесь нужен рисунок - функция Шекеля, и два варианта регрессии - полином и деревья решений. Шекель должен плохо полиномами представляться, т.к. там функции вида 1/x
Одновременно с этим регрессия на основе дерева решений позволяет с достаточной точностью легко выделять области притяжения локальных экстремумов.

Построение регрессии с помощью дерева решений включает два основных этапа:
\begin{itemize}
	\item Область поиска $D$ разбивается на $J$ непересекающихся подобластей $D_1, D_2, ..., D_J$, причем $D = \bigcup_{j=1}^{J}{D_j}$.
	\item Любому значению, попадающему в область $D_j$, т.е. $x \in D_j$, ставится в соответствие среднее значение $c_j$ по значениям training trials, попавших в эту область.
\end{itemize}

Фактически, деревья решений строят модель функции вида
\begin{equation}\label{dtree}
f(x) = c_j, \; x\in D_j.
\end{equation}

Указанная модель является, с одной стороны, довольно простой, а с другой стороны, адекватно отражает интересующие нас свойства функции (наличие или отсутствие локальных минимумов). Выделение областей притяжения локальных минимумов с использованием модели (\ref{dtree}) можно организовать следующим образом.

Пусть $x^{k+1}$ -- точка текущего испытания. Для данной точки ищется номер $j$, такой что $x^{k+1} \in D_j$. 
Далее сравниваются значения $c_j$, соответствующие соседним подобластям. Если выполняется одно из условий
\begin{gather}\label{LocalPointC}
c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3} \leq c_{j+4}, \; j=1; \nonumber \\ 
c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3}, \; j=2;  \nonumber \\  
c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2}, \; 3 \leq j \leq J-2; \\ 
c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1}, \; j=J-1; \nonumber \\ 
c_{j-4} \geq c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j}, \; j=J; \nonumber 
\end{gather}
то область $D_j$ считается областью притяжения локального минимума. В ней можно запускать локальный поиск, а впоследствии -- исключить данную подобласть из глобального поиска.
В качестве локального алгоритма может использоваться любой из локальных методов нулевого порядка, например,  golden section search or parabolic interpolation \cite{Press}.

Для того, чтобы модифицировать алгоритм глобального поиска из subsection \ref{CoreGSA} с целью исключения областей притяжения локальных минимумов, сопоставим каждой точке испытания $x^i$, получаемой при работе алгоритма, дополнительный признак $q^i \in \{0,1,2\}$, который будет характеризовать свойства данной точки. 
Значение $q^i=0$ присваивается по умолчанию и означает что точка $x^i$ получена в результате работы правила (\ref{xk1}) алгоритма глобального поиска.
Значение $q^i=1$ присваивается точкам, полученным в результате работы локального метода, при этом значение $q^i=2$ соответствует точке локального минимума, найденной в результате работы локального поиска.

Напомним, что верхний индекс соответствует номеру итерации, на которой было проведено испытание в данной точке, а нижний индекс -- номеру точки в ряду (\ref{xt}).

Опишем теперь модифицированный алгоритм глобального поиска, использующий деревья решений для выделения и исключения областей притяжения локальных минимумов; будем в дальнейшем ссылаться на данный алгоритм как на GSA-DT.

Шаги 1--5 алгоритма GSA-DT полностью повторяют шаги 1--5 .

Step 6. Если $q_{t-1} \in \{1,2\}$ или $q_t \in \{1,2\}$, то перейти к проверке критерия остановки. Иначе перейти на Step 7.

Step 7. Построить дерево решений на основе результатов выполненных испытаний, и получить соответствующую кусочно линейную аппроксимацию, которая каждой подобласти $D_1, D_2, ..., D_J$ сопоставляет значение $c_1, c_2, ..., c_J$.

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning}
		   %Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Department of Mathematical Software and Supercomputing Technologies, Lobachevsky University, 603950 Nizhny Novgorod, Russia; ilya.lebedev@itmm.unn.ru (I.L.)}
%$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
The article features the study of global optimization problems and numerical methods of their solution. Such tasks are computationally expensive since the objective function can be multi-extremal, nondifferentiable, and, as a rule, given in the form of a "black box". The study used a deterministic algorithm for finding the global extremum. The said algorithm is based neither on the concept of multistart, nor nature-inspired algorithms. The article provides computational rules of the one-dimensional algorithm and the nested optimization scheme which could be applied for solving complex multidimensional problems. Note that the solution complexity of global optimization problems essentially depends on the presence of multiple local extrema. In this paper, we apply machine learning methods to identify areas of attraction of local minima. The use of local optimization algorithms in the selected areas can significantly accelerate the convergence of global search as it could reduce the number of search trials in the vicinity of local minima. The results of computational experiments carried out on several hundred global optimization problems of different dimensionalities presented in the paper confirm the effect of accelerated convergence (in terms of the number of search trials required to solve a problem with a given accuracy).
}

@@

The successful application of machine learning (ML) methods to solve a wide range of problems leads to the emergence of new ways to apply ML for many tasks. 
Methods of machine learning were shown to be particularly effective for identifying the principal properties of the phenomena (for example, physical, economic, or social), which are stochastic by nature or contain some hidden parameters \cite{Golovenkin2020,Gonoskov2019}.
ML is also successfully used to solve complex problems of computational mathematics, for example, for simulation of dynamical systems  \cite{Seleznev2019}, solution of ordinary, partial, or stochastic differential equations \cite{Lagaris1998,Blechschmidt2021,Xu2020} .

For example, ML could be applied for solving such a complex problem of computational mathematics as global optimization.
The solution to this class of problems, as a rule, cannnot be found analytically and, therefore, one needs to construct numerical methods to solve it.

The numerical solution of optimization problems is fraught with significant difficulties. In many ways, they are related to the dimensionality and type of the objective function. Consequently, the most difficult tasks are those in which the objective function is multi-extremal, nondifferentiable, and, moreover, given in the form of a black box (i.e., in the form of some computational procedure, the input of which is an argument, and the output is the corresponding value of the function). These complex problems are the main focus of this article.

There are several approaches to the construction of numerical methods for solving global optimization problems. 
%Надо ли добавить алгоритмы на основе метамоделей ?
Some algorithms are based on the idea of a multistart: launching a local search either from different starting points or with varying parameters. Local optimization methods have a high convergence rate. At the same time, one of the main problems in multistart schemes is the choice of starting points that would correspond to the regions of attraction of various local solutions. 
Machine learning methods can be successfully applied to solve this problem.  
For example, in \cite{RinnooyKan1987} methods of cluster analysis were used to select promising starting points
In \cite{Cassioli2012} ], the area for launching the local method was allocated based on the classification of starting points using a support vector machine.

Another popular class of methods for solving global optimization problems is metaheuristic algorithms. 
Many of them are based on imitation of the processes occurring in living nature. The parameters of such algorithms could also be tuned using ML. For example, \cite{Jin2005} provides an overview of machine learning applications in evolutionary algorithms.

Note that the algorithms of the last two classes do not provide guaranteed convergence to the solution of the problem and are inferior to deterministic algorithms in terms of the quality of solution \cite{Kvasov2018,Sergeyev2018} ] (eg, measured by the number of correctly solved problems from a particular set). Therefore, deterministic methods seem to be potentially more effective.

@@

Several strategies have been proposed to speed up the operation of the global search algorithm (in terms of the number of iterations required to solve the problem with a given accuracy). In this paper, we propose a new approach to acceleration based on identifying areas of attraction of local minima using machine learning methods. The identification of regions of attraction and the launch of local search in these regions can significantly reduce the number of trials required for the method to achieve global convergence. Experiments carried out on a series of several hundred test problems confirm this statement.

@@

In this paper, we will consider global optimization problems of the form
\begin{eqnarray}\label{main_problem}
& \varphi(y^\ast)=\min{\left\{\varphi(y):y\in D\right\}},\\
& D=\left\{y\in \text{R}^N: a_i\leq y_i \leq b_i, 1\leq i \leq N\right\}. \nonumber
\end{eqnarray}
Problem (\ref{main_problem}) is considered under the assumption that the objective function is multi-extreme, is given in a form of a "black box" , and the calculation of its values is associated with solving the problem of numerical modeling, which makes the solution a labor-intensive operation.

A typical situation for many applied problems is when a limited change in the vector of parameters y causes a limited change in the values of $\varphi(y)$. The mathematical model describing this premise is based on the assumption that the Lipschitz condition is satisfied
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty.
\]
The common assumption for many approaches to the development of optimization algorithms is that the objective function is Lipschitzian  \cite{Jones1993,Pinter1996,Zilinskas2008,Evtushenko2009}.
At the same time, many known approaches are based on various methods of dividing the search domain into a system of subdomains and then choosing the most promising subdomain for placing the next trial (calculating the value of the objective function)  \cite{Jones2009,Zilinskas2010,Evtushenko2013,Kvasov2013,Paulavicius2016}. 
An important property of global optimization problems is the fact that, in contrast to the problems of finding a local extremum, the global minimum is an integral characteristic of the problem being solved. Making sure that the point  $y^*\in D$ is a solution to the problem requires going beyond its neighborhood to the investigation of the entire search domain. As a result, when minimizing substantially multi-extremal functions, the numerical method must construct a coverage of the search domain. As a result, the number of nodes of this coverage increases exponentially with increasing dimensionality. 
This feature determines the high complexity of solving multiextremal optimization problems making dimensionality a critical factor that affects the complexity of their solution.

The dimensionality in multi-extreme optimization leads to many problems, so, scholars use a wide variety of approaches to reducing it.  
For example, simplicial or diagonal partition of the search domain allows using methods for solving one-dimensional problems to solve the original multidimensional problem (see, for example,  \cite{PaulaviciusZilinskas2014,Sergeyev2017,Sergeyev2013}). 
Another well-known approach to dimensionality reduction is using the Peano space-filling curves to map the multidimensional domain onto a one-dimensional interval \cite{Strongin2000}.

In this work, we will use another method based on the nested optimization scheme \cite{Shi2000,Grishagin2001,VanDam2010,Grishagin2015} and its generalization \cite{Grishagin2016,Grishagin2016_1}.
The nested optimization scheme, on the one hand, does not worsen the properties of the minimized function (unlike reduction using Peano curves), and, on the other hand, does not require the use of complex data structures to support simplex or diagonal partitions of the feasible region
At the same time, the nested optimization scheme makes it possible to reduce the solution of a multidimensional problem to the solution of a set of standard one-dimensional problems, which, then, can be solved by a wide range of one-dimensional global optimization algorithms.

@@

As a standard, let us consider a one-dimensional multiextremal optimization problem
\begin{equation}\label{uni_problem}
\varphi^\ast = \varphi(x^\ast)=\min{\left\{\varphi(x):x\in \left[0,1\right] 
\right\}}
\end{equation}
with an objective function satisfying the Lipschitz condition

Here is the description of global search algorithm (GSA) for solving the basic problem in accordance with \cite{Strongin2000}.
]. In the course of its work, GSA generates a sequence of points  $x^i$, at which the values of the objective function  $z^i=\varphi(x^i)$ are calculated. 
We will refer to the process of calculating the value of the objective function as  \textit{trial}.

In accordance with the algorithm, the first two trials are carried out at the boundary points of the segment $[0,1]$, т.е. $x^0=0,\;x^1=1$. 
At these points, the values of the objective function  $z^0=\varphi(x^0),\;z^1=\varphi(x^1)$ и уare calculated and the counter value is set to $k=1$. 
The point of the next trial $x^{k+1}, k\geq 1,$ is selected in accordance with the following procedure:

 Step 1. Reindex  (starting at 0) the points xl, $x^i,\:0\leq i\leq k$, of the trials conducted  in ascending order of the coordinate, i.e.
\begin{equation}\label{xt}
0=x_0<x_1<\ldots <x_{k}=1.
\end{equation} 
Compare the values of the objective function $z_i=\varphi(x_i), \; 0\leq i\leq k$ to the points $x_i, \; 0\leq i\leq k$, at which these values were calculated.

Step 2. Calculate the maximum absolute value of the relative first difference
\begin{equation}\label{mu}
\mu=\max_{1\leq i\leq k}\frac{\left|z_i-z_{i-1}\right|}{\Delta_i},
\end{equation}
where $\Delta_i = x_i-x_{i-1}$. If the value calculated in accordance with this formula is equal to zero, then take $\mu = 1$.

Step 3. For all the intervals $(x_{i-1},x_i),1\leq i\leq k$,  calculate the value
\begin{equation}\label{R}
R(i)=r\mu\Delta_i+\frac{(z_i-z_{i-1})^2}{r\mu\Delta_i}-2(z_i+z_{i-1}),
\end{equation} 
refered to as the \textit{characteristics} of the interval; value $r>1$ is the parameter of the algorithm.

Step 4. Find interval $(x_{t-1},x_t)$ with the maximum characteristics
\begin{equation}\label{MaxR}
R(t)=\max_{1\leq i\leq {k}}R(i).
\end{equation}
If the maximum characteristics corresponds to several intervals, then choose the minimum number that satisfies as $t$ (\ref{MaxR}).

Step 5. Carry out a new trial at the point
\begin{equation}\label{xk1}
x^{k+1}=\frac{1}{2}(x_{t-1}+x_t) - \frac{z_t-z_{t-1}}{2r\mu}.
\end{equation}

The algorithm stops when the condition  $\Delta_t<\epsilon$ is satisfied; here $t$ from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy. 
The following values are selected to evaluate the efficiency of the solution
\[
z_k^\ast=\min_{0\leq i \leq k}\varphi(x^i), \ x_k^\ast=\arg \min_{0\leq i \leq
 k}\varphi(x^i).
\]

The theoretical conditions that determine the convergence of the algorithm are presented in \cite{Strongin2000}. 
The operation of the algorithm during the minimization of a specific multiextremal function, which is specified in accordance with formula  (\ref{hill}), is shown in Figure \ref{fig1}. The algorithm was launched with the parameter $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$in the criterion for stopping the method.
Figure \ref{fig1} shows the objective function graph and points of 71 search trials which GSA needed to solve the problem to the specified accuracy. This highlights the problem of all methods of global optimization - the concentration of points in the vicinity of local minima of the problem, which are not a global solution.

\begin{figure}[H]
\includegraphics[width=1.0\linewidth]{HillAGP90.png}
\caption{Minimization of a function of the form  (\ref{hill}) using GSA.}
\label{fig1}
\end{figure}

\subsection{Machine Learning Regression as a tool for identifying areas of attraction of local extrema}\label{TreeGSA}

The functions considered in this study belong to the class of Lipschitzian functions. Therefore, classical regression methods (for example, polynomial regression, where a function is approximated by a polynomial of a given degree) will not properly match the behavior of the function.
A more powerful tool for this task is regression splines. When constructing a regression spline, the domain of definition is divided into K non-overlapping subdomains. In each of such domains, the function is approximated by a polynomial. Dividing the interval into a sufficient number of subdomains allows one to very accurately approximate the original function.

Regression can also be constructed using such a powerful tool as artificial neural networks. 
Different types of networks can be used to build the regression, for example, multilayer perceptron, radial basis function network, etc.
However, in both of these cases, the model itself (spline or neural network) becomes rather complex for the analysis required to solve the given problem (identifying areas of attraction of local extrema).

Therefore, within the framework of the study, we chose a regression model based on decision trees to analyze the local behavior of a function.
For example, if the objective function is properly approximated by polynomials, then the polynomial regression, of course, will appropriately convey the properties of the function. However, if there is a more complex relationship between them, then the decision tree can surpass the classical variants of regression in terms of the quality of the approximation. 
%Здесь нужен рисунок - функция Шекеля, и два варианта регрессии - полином и деревья решений. Шекель должен плохо полиномами представляться, т.к. там функции вида 1/x
At the same time, the regression based on the decision tree makes it possible to easily identify the areas of attraction of local extrema with sufficient accuracy.

Building a regression using a decision tree consists of two main steps:
\begin{itemize}
	\item Search domain $D$ is divided into $J$ non-overlapping subdomains  $D_1, D_2, ..., D_J$, provided that $D = \bigcup_{j=1}^{J}{D_j}$.
	\item Any value falling into the subdomain $D_j$, i.e. $x \in D_j$, is matched to the average value  $c_j$ based on the training trial values that fall into this subdomain. 
\end{itemize}

In fact, decision trees build a model of a function of the form
\begin{equation}\label{dtree}
f(x) = c_j, \; x\in D_j.
\end{equation}

This model is, on the one hand, rather simple, and on the other hand, it adequately reflects the properties of the function under investigation (the presence or absence of local minima). The selection of regions of attraction of local minima using model (\ref{dtree}) can be organized as follows.

Let $x^{k+1}$ be the point of the current trial. For a given point, an index  $j$ is sought such that $x^{k+1} \in D_j$. 
Next, the  $c_j$ values corresponding to neighboring subdomains are compared. If one of the conditions is met
\begin{gather}\label{LocalPointC}
c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3} \leq c_{j+4}, \; j=1; \nonumber \\ 
c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2} \leq c_{j+3}, \; j=2;  \nonumber \\  
c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1} \leq c_{j+2}, \; 3 \leq j \leq J-2; \\ 
c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j} \leq c_{j+1}, \; j=J-1; \nonumber \\ 
c_{j-4} \geq c_{j-3} \geq c_{j-2} \geq c_{j-1} \geq c_{j}, \; j=J; \nonumber 
\end{gather}
then the subdomain $D_j$ is considered the area of attraction of the local minimum. Here, one can start a local search, and subsequently exclude this subdomain from the global search. 
Any of the zero-order local methods can be used as a local algorithm, for example, golden section search or parabolic interpolation \cite{Press}.

In order to modify the global search algorithm from subsection  \ref{CoreGSA} to exclude the regions of attraction of local minima, we associate each test point  $x^i$, obtained during the operation of the algorithm with an additional criterion $q^i \in \{0,1,2\}$, which will characterize the properties of this point. 
The value  $q^i=0$ is assigned by default and indicates that point $x^i$ is obtained as a result of rule (\ref{xk1}) of the global search algorithm.
The value $q^i=1$ is assigned to the points obtained as a result of the work of the local method, while the value  $q^i=2$ corresponds to the point of the local minimum found as a result of the work of the local search.

Recall that the superscript corresponds to the number of the iteration at which the trial was carried out at this point, and the subscript corresponds to the number of the point in the row  (\ref{xt}).

Let us now describe a modified global search algorithm that uses decision trees to isolate and exclude areas of attraction of local minima; we will further refer to this algorithm as GSA-DT.

Steps 1-5 of the GSA-DT algorithm are the same as steps 1-5.

Step 6. If $q_{t-1} \in \{1,2\}$ or $q_t \in \{1,2\}$, then go to check the stopping criterion. Otherwise, go to Step 7.

Step 7. Construct a decision tree based on the results of the tests performed, and obtain the corresponding piecewise linear approximation, which assigns the value $c_1, c_2, ..., c_J$ to each subdomain  $D_1, D_2, ..., D_J$.

Step 8. For the point $x^{k+1}$ of the current trial, find a number $j$ such that $x^{k+1} \in D_j$ and check whether the condition  (\ref{LocalPointC}) is satisfied.If condition (\ref{LocalPointC}) is satisfied, start a local search in the domain $D_j$ from the point  $x^{k+1}$. 
The results of all trials performed during the local search at the points $x^{k+2}, ...,x^{k+k_{local}}$ are stored in the information base of the algorithm and are used at subsequent iterations. 
All these points receive the attribute  $q^i=1$, $i = k+2, ... , k+k_{local}$. The attribute equal to 2 is assigned to the point corresponding to the found local minimum.

The stopping criterion of the modified algorithm will look as follows. 
The algorithm stops when one of the following conditions is met:

\begin{itemize}
	\item $|x_{t} - x_{t-1}|<\epsilon$, where $t$ from (\ref{MaxR}), and $\epsilon>0$ is a given accuracy;
	\item $|x^{k+1} - x_{t-1}|<\epsilon$ and $q_{t-1} = 2$,
	\item $|x^{k+1}-x_{t}|<\epsilon$ and $q_{t} = 2$.
\end{itemize}

For example, consider the operation of the GSA-DT algorithm with minimization of the same multi-extreme function, which is presented  Figure \ref{fig1}. The same parameters were used at the start of the algorithm: the parameter  $r=2.2$ from (\ref{R}) and value $\epsilon = 10^{-3}$ in the stopping criterion of the method. 
Figure \ref{fig2} illustrates the operation of the GSA-DT algorithm. In addition to the graph of the objective function, it shows a piecewise constant approximation of the form (\ref{dtree}) built at the final stage of the search. Black points on the graph correspond to the global search phase, green points correspond to the work of the local method. In total, the GSA-DT method required 49 trials to solve the problem; there was no accumulation of trial points in the vicinity of local minima.
